The powder x-ray diffractogram open database

### 2024-04-28 paper/03_our_dataset.tex
New title idea
Moved main tex file and references to repo root

    \subsection{Synthetic distribution} \label{sec:synth_distribution}

    \begin{figure}[!htb]
    \centering
    \includegraphics{./figures/example_structures.pdf}
    \caption{a) Some \replaced{randomly chosen and thus representative}{randomly picked} examples of ICSD crystals. b) Some \replaced{randomly chosen and thus representative}{randomly
    picked} examples of synthetically generated crystals. While coordination and
    distances are not chemically correct for the synthetic crystals, crystal
    symmetries are reproduced correctly.}
    \label{fig:example_structures}
    \end{figure}

    \begin{figure}[!htb]
    \centering
    \includegraphics{./figures/histograms/histograms.png} 
    \caption{Histograms comparing the distributions of descriptors of the
    synthetically generated crystals with the ICSD distribution in the test
    dataset. a) density factor $ \frac{V_\text{unit cell}}{\sum_i 4/3 \pi
    \left(\frac{r_{i\text{;cov}} + r_{i\text{;VdW}}}{2}\right)^3} =
    \frac{V_\text{unit cell}}{V_\text{atomic}}$, b) crystallite sizes, c) unit cell
    volume (conventional cell settings), d) number of atoms in the asymmetric
    unit. The probability density of the ICSD is visualized by a stacked bar
    histogram, where the green portion of the bar was correctly classified and
    the red portion was incorrectly classified. The probability density of the
    synthetic crystals is visualized by the dark blue line. The portion between the
    dark blue line and the light blue line was correctly classified, the portion below the
    light blue line was incorrectly classified. The reported classification
    performance is based on the ResNet-101 model trained on diffractograms from
    synthetic crystals.}
    \label{fig:histograms}
    \end{figure}

    We first present an analysis of the generated synthetic crystals.
    Figure~\ref{fig:example_structures} shows some \replaced{randomly chosen and thus representative examples of}{randomly selected} ICSD and
    synthetic crystal structures side-by-side. Visually, the crystals appear very
    similar. However, no physical or chemical considerations regarding stability, clashing atoms, and element combinations are taken into account in the generation of synthetic crystals. As discussed earlier, our goal is to demonstrate that this is
    not problematic when using these crystals for the extraction of structural
    information from powder diffractograms. On the contrary, we expect the synthetic crystals to be a better basis for generalization to fundamentally new crystal structures than existing finite databases.

    To compare the distribution of ICSD crystals with the synthetic
    distribution, we evaluated structural descriptors, i.e. density factors, crystallite sizes, unit cell volumes, and numbers of atoms in the asymmetric unit, and compare their histograms
    in Figure~\ref{fig:histograms}.
    One can see that the overall distributions of
    the synthetic and ICSD crystals are very similar for all four descriptors.
    This shows that our chosen generation algorithm reproduces crystals that are
    similar to ICSD crystals in terms of these more general descriptors.

    \subsection{Classification results} \label{sec:classification_results}

    The main results of our experiments (see Section~\ref{sec:models_and_experiments}) to
    classify the space group of powder diffractograms can be
    found in Table~\ref{tab:results}. \added{In SI Table S2, we further provide the training time and total number of unique diffractograms for each computational experiment.}
    The goal of our experiments is to systematically analyse and quantify the changes in classification accuracy introduced by our two main contributions: A more challenging dataset split, and training on continuously generated synthetic data.

    We started by repeating previously reported
    experiments\supercite{parkClassificationCrystalStructure2017} trained
    directly using ICSD crystals with a random train-test split instead of the
    split based on structure types. This model achieved a very high test
    accuracy of 83.2\%. We note that the previous publication that we compare
    our results to\supercite{parkClassificationCrystalStructure2017} removed
    data from the training dataset, ``[...] heavily duplicated data
    [...]''\supercite{parkClassificationCrystalStructure2017}, but did not
    specify the exact criterions used. In contrast, we did not exclude any
    duplicates in this experiment based on a random train-test split.
    Furthermore, as discussed in Section~\ref{sec:generate_crystals}, we
    excluded crystals with a very high unit cell volume and a very high number
    of atoms in the asymmetric unit. This is likely the reason for the slightly
    higher classification accuracy that we observed, compared to the originally
    reported 81.1\%.

    When splitting randomly, the model merely needs to recognize structures or
    structure types and assign the correct space group. This task is much easier
    than actually extracting the space group using more general patterns. When
    going from random splits to structure type-based splits (see
    Section~\ref{sec:dataset_split}), it becomes obvious that both the ``parkCNN
    big'' as well as the ``parkCNN medium'' models overfit the training data and
    do not generalize well to unseen structure types in the test set \added{(see
    Table \ref{tab:results})}. \added{The ``parkCNN medium'' model, which
    achieved 83.2\% on a random split, now only yields 55.9\% with the structure
    type-based split.}

    \replaced{Training the models by
    \citeauthor{parkClassificationCrystalStructure2017}}{Training the same
    model}, in particular the ``parkCNN big'' model, on synthetic crystals leads
    to a 1.6 percentage points higher test accuracy than the ``parkCNN big''
    model trained on ICSD diffractograms. At the same time, the training
    accuracy drops from \added{the} 87.2\% \added{when we trained the model
    directly on the ICSD} to 74.2\% \added{on the synthetic distribution}
    indicating that the model is now limited more by missing capacity rather
    than by overfitting, which is why we explored larger models, which will be
    discussed later. The gap between training and test accuracy is $31.1$
    percentage points when training on ICSD data, while for training using
    synthetic crystals, the gap is only $16.5$ percentage points. We note that
    this gap between training using synthetic crystals and testing using ICSD
    crystals cannot stem from overfitting, since no diffractograms are repeated
    for the synthetic training. The difference rather stems from the differences
    between the synthetic distribution and the ICSD distribution of crystals.

    \begin{table}[!htb]
        \begin{center}
        \caption{Results of training on diffractograms simulated from ICSD crystals (random splits as well as structure type-based splits) compared to when training on diffractograms from synthetic
        crystals. Test accuracy
        in all cases refers to the accuracy when testing on the ICSD test
        dataset. The training accuracies are averaged over the last 10 epochs of
        the respective run.
        Experiments trained directly on ICSD data
        overfitted to the training data. Training longer would have further
        increased the training accuracy, while not increasing the test
        accuracy.}
        \vspace*{2mm}

        \begin{tabular}{ccccccc} 
            \toprule
            Split & \begin{tabular}{@{}c@{}}Training \\ dataset\end{tabular} & \begin{tabular}{@{}c@{}}Testing \\ dataset\end{tabular} & Model & \begin{tabular}{@{}c@{}}Number of \\ parameters\end{tabular}& \begin{tabular}{@{}c@{}}Training acc. \\ / \%\end{tabular} & \begin{tabular}{@{}c@{}}Test acc. \\ / \%\end{tabular} \\
            \midrule
            Random & ICSD & ICSD & parkCNN medium & 4\,246\,797 & 88.4 & 83.2 \\
            \midrule
            Structure& \multirow{2}{*}{ICSD} & \multirow{2}{*}{ICSD}& parkCNN big & 4\,959\,585 & 87.2 & 56.1 \\
            type&&& parkCNN medium & 4\,246\,797 & 90.9 & 55.9 \\
            \midrule
            & \multirow{4}{*}{synthetic} & \multirow{4}{*}{ICSD} &parkCNN big & 4\,959\,585 & 74.2 & 57.7 \\
            Structure&&& ResNet-10 & 9\,395\,025 & 87.2 & 73.4 \\
            type$^{1}$ &&& ResNet-50 & 41\,362\,385 & 91.8 & 79.3 \\
            &&& ResNet-101 & 60\,354\,513 & 92.2 & \textbf{79.9} \\
            \bottomrule
        \end{tabular}\label{tab:results}
        \end{center} 
        \footnotesize{$^{1}$Here, the split type refers to the statistics and the test dataset, rather than the training and the test dataset.}
    \end{table}

    While the ``parkCNN big'' model trained on synthetic crystals outperforms
    the approach of training directly on ICSD crystals by only 1.6 percentage
    points, the advantage of training on an infinite stream of synthetic data increases when using models with more parameters and thus higher capacity. In
    contrast to training directly on a finite set of ICSD crystals, it is possible to train
    very large models using the infinite synthetic data stream without the
    potential of overfitting. As found in the last lines of Table~\ref{tab:results}, ResNet-10, ResNet-50, and ResNet-101 based models achieve ICSD test accuracies of 73.4\%,
    79.3\%, and 79.9\%. This is a significant increase from the 57.7\% achieved
    by the ``parkCNN big'' model. Figure~S4 in the SI further shows the top-$k$ accuracy over $k$ for the ResNet-101 model. With increasing $k$ the accuracy exceeds 95\% at $k=5$.
    This means that our model can not only determine the correct space group with a high probability but can also generate an almost complete list of possible space group candidates.

    Figure~\ref{fig:training_curve} shows the ICSD test accuracy, the training
    accuracy (on synthetic data), and the ICSD top-5 test accuracy for all three ResNet variants
    as a function of epochs trained. For all three metrics, the difference
    between ResNet-50 and ResNet-101 is comparably small, while the step from
    ResNet-10 to ResNet-50 is substantial (5.9 percentage points in ICSD test
    accuracy, see Table~\ref{tab:results}). This shows that going beyond the
    model size of the ResNet-101 will likely not yield a big improvement in
    accuracy. In contrast to the 79.9\% accuracy reached in the top-1 ICSD test accuracy, the top-5 ICSD test accuracy of the ResNet-101 model reaches
    96\%. However, for all three ResNet variants, a gap between training using synthetic
    crystals and testing using the ICSD remains (12.3 percentage points for
    ResNet-101). As also shown in Figure~S5 in the SI, the accuracy convergence can be approximately described by a power law, indicating that exponentially more training time will substantially reduce classification errors and thus potentially lead to top-1 accuracies of 90\% and above, at the cost of a 100-fold increase in training times\replaced{. Considering the current training times provided in Table S2 of the SI, this is currently infeasible or only possible with tremendous hardware resources.}{, which is currently infeasible.}

    \begin{figure}[!htbp] 
    \centering
    \includegraphics{figures/training_curves/training_curve_main.png}
    \caption{Test accuracy (ICSD), training accuracy (synthetic crystals), and
    test top-5 accuracy (ICSD) as a function of epochs\added{ (bottom axis). Since each additional epoch contains newly generated unique diffractograms, we further show the accuracies as a function of the total number of unique synthetic diffractograms (top axis)}. We show all three
    metrics for the models ResNet-101, ResNet-50, and ResNet-10. To better show
    the scaling behavior, both axes use logarithmic scaling. \added{Figure S6
    shows the same plot but without logarithmic scaling.} To better see the
    exponential behaviour, see Figure~S5 in the SI. }
    \label{fig:training_curve} 
    \end{figure}

    The histograms in Figure~\ref{fig:histograms} show, next to the overall
    distribution, also the fraction of diffractograms classified wrongly for
    testing on the ICSD (red bar) and on the synthetic data (below the light blue line)
    for the ResNet-101 model. First, one can see that throughout almost all
    regions of the distributions, the accuracy on the synthetic data is slightly
    higher than that on the ICSD. This is related to the aforementioned gap of 12.3
    percentage points between train and test accuracy and can be attributed to
    differences between the synthetic and ICSD distribution of crystals. This
    will be discussed in detail in the next section. It is surprising to see that the
    dependence on crystallite sizes is rather weak, as smaller crystallite sizes
    result in broader peaks (see Scherrer equation, Eq.~\ref{eq:scherrer}), potentially making the
    classification harder due to more peak overlaps.

    In summary, the maximum ICSD test accuracy of 79.9\% that we achieved using
    the ResNet-101 model almost reaches the previously
    reported\supercite{parkClassificationCrystalStructure2017} 81.14\% for the
    space group classification. However, our accuracy is based on a train-test
    split based on structure types, in contrast to a random split. This creates
    a much harder but also realistic task to solve since the model needs to
    generalize to other structure types without merely recognizing
    diffractograms or structure types that it has already seen during training.
    This becomes especially apparent from our experiment directly trained on
    diffractograms from ICSD crystals with the split based on structure types,
    which reached only 56.1\% instead of the previously
    reported\supercite{parkClassificationCrystalStructure2017} 81.14\%.

    \subsubsection*{Experimental results}
    To go beyond simulated diffractograms, we trained ResNet-50 models on
    calculated diffractograms with background, noise, and impurities and applied
    the trained models to the RRUFF mineral database. Our results (see Figure~S3
    in the SI) show that it is essential to include impurity phases in the
    training data. By doing so, we obtain a top-1 accuracy of 25.2\% and a
    top-10 accuracy of over 60\%. This is of high practical relevance since
    having a short list of potential space groups is often sufficient as a first
    step to further refinement and analysis.

    \citeauthor{vecseiNeuralNetworkBased2019} performed similar experiments of
    space group classification on the same database. Using an ensemble of 10
    fully connected neural networks, they reached a classification accuracy of
    54\%\supercite{vecseiNeuralNetworkBased2019}. While our obtained accuracy is
    significantly lower, our approach is much more general: In contrast to our
    approach, the training dataset was based on simulated diffractograms of
    structures of the ICSD\supercite{vecseiNeuralNetworkBased2019}, which
    contains almost all RRUFF structures, leading to high similarities of
    training and \replaced{test}{testing} data. Therefore, the model needed to
    simply recognize the minerals, instead of directly inferring the space group
    using the symmetry elements - as our method needs to do.

    We want to emphasize that our efforts to apply the methodology to
    experimental data are only preliminary. We expect improved results with an
    improved data generation protocol since the procedure contains many
    parameters to be tuned. Ideally, one would use a generative machine learning
    approach to add the experimental effects (noise, background, impurities) to
    the pure diffractograms. We also want to point out that the noise level and
    quality of data in the RRUFF dataset are limited. Application of the
    presented methodology \replaced{to}{on} other experimental datasets is
    desirable. \deleted{We will address these points in future work, where we
    focus on improved ways of modeling experimental imperfections.}
    \added{As discussed above, for the
    classification of pure diffractograms we observed the ResNet-50 to have the
    best cost-benefit ratio, since the ResNet-101 yielded only slight
    improvements. For the more complicated problem of classifying diffractograms
    with experimental imperfections, bigger models and longer training times
    might be necessary.}

    \added{Next to improving the modeling of experimental imperfections and therefore the overall accuracy on experimental data, the practical application of deep neural networks for analyzing powder diffractograms yields further challenges that we want to discuss. Since experimental setups differ, e.g., concerning the used wavelength, a different $2\theta$ step size, or a different $2\theta$ range, a new neural network would need to be trained for each situation. Since our largest model requires a significant computational investment, this might not be feasible in all situations. Arguably, though, for large high-throughput experiments, the 11-day training of a ResNet-50 should not be unreasonable, especially if it can speed up the data analysis significantly and allow in-loop adaptive experimentation. For smaller setups, where this is not feasible, other solutions must be found. First, one can use a form of transfer learning from a pre-trained model to fine-tune to the desired experimental setup. This, however, would only work for a change in wavelength, since a change in step size or $2\theta$ range would change the input dimensions of the network. However, to handle a change in the $2\theta$ range, it might be possible to include a form of zero-masking in the synthetic training data, such that different input ranges (with zeros where no measurement was made) can be used, which would lead to a more flexible model, not requiring new training data when applied to a new $2\theta$ range. For a change in the step size, a cubic spline interpolation might be helpful. We plan to address these challenges in future work.}

    \deleted{As discussed above, for the classification of pure diffractograms we
    observed the ResNet-50 to have the best cost-benefit ratio, since the
    ResNet-101 yielded only slight improvements. For the more complicated
    problem of classifying diffractograms with experimental imperfections,
    bigger models and longer training times might be necessary.}

    \added{Furthermore, analysis of the loss value or gradient norm associated with particular samples, i.e. crystal structures, during training on synthesis crystals or during transfer learning from synthetic to experimental data can help to better understand the relevance and informativeness of given samples for the model. This can help in generating more relevant synthetic data based on experimental crystal structures that are underrepresented in the synthetic data distribution.}

    \subsection{Differences between synthetic crystals and ICSD crystals}
    We showed that training directly on crystals from the ICSD yields a gap
    between the training and test accuracy due to overfitting. The training on
    the synthetic dataset also shows a gap between the training and test
    accuracy (see Table~\ref{tab:results}), but it is smaller than when training
    directly on ICSD crystals. Furthermore, this gap is not due to overfitting,
    since overfitting to singular diffractograms is not possible when the model
    is trained using an infinite stream of generated synthetic crystals. The gap
    rather stems from systematic differences between the synthetic and ICSD
    distribution of crystals.

    To analyze those differences, we created three modifications of the ICSD
    test dataset \added{(see SI Section~S3 for details)}. In the first modification, the fractional coordinates of the
    atoms in the asymmetric unit of the crystals of the ICSD test dataset were
    randomly uniformly resampled (as in the synthetic crystal generation
    algorithm). In the second modification, the lattice parameters were
    randomized following the \replaced{kernel density estimate}{KDE} used in the synthetic generation algorithm. The
    third modification combines both previous modifications, i.e. both the
    coordinates and the lattice parameters were resampled. These three modified
    test datasets bring the ICSD test dataset closer to the distribution used
    for training and let us quantify which factors contribute to the gap between
    training \added{on synthetic crystals} and testing \added{on the ICSD}.

    We evaluated the test accuracies on \replaced{the randomized}{these}
    datasets for the experiment using the ResNet-101 model trained using
    synthetic crystals. We found that randomizing the coordinates yields an
    increase in test accuracy of \replaced{$4.89$}{$5.96$} percentage points.
    Randomizing the lattice parameters results in an increase of
    \replaced{$0.79$}{$1.31$} percentage points. Randomizing both the
    coordinates and the lattice parameters leads to an increase of
    \replaced{$5.70$}{$6.32$} percentage points, explaining
    \replaced{almost}{approximately} half of the gap of 12.3 percentage points
    between synthetic training and ICSD test accuracy.

    \added{So far, we have randomized the lattice parameters and coordinates of the test dataset, such that they follow a distribution that is based on the statistics extracted from the statistics dataset. However, this does not take into account the different Wyckoff position occupation probabilities between the test and statistics datasets. For this, we repeated
    a similar analysis, for which we applied the
    randomizations to the statistics dataset rather than the test dataset.
    Without any modifications, testing on the statistics dataset instead of the test dataset yielded $3.89$ percentage points higher accuracy. This can be explained by slight differences in the overall statistics between the test and statistics datasets.
    Randomizing the coordinates yields a further increase of $4.72$
    percentage points, randomizing the lattice $1.16$ percentage points, and
    randomizing both the coordinates and the lattice parameters $6.68$ percentage points.
    In total, testing on the statistics dataset with randomized coordinates and
    lattice parameters yields a $10.57$ percentage points higher accuracy than on the
    unmodified test dataset. This almost completely explains the gap
    of $12.3$ percentage points between the training accuracy on synthetic crystals and the test accuracy on the ICSD. The
    remaining part is likely due to our algorithm that places atoms on Wyckoff
    positions 
    not reproducing the ICSD distribution exactly. 
    However, the
    remaining difference is remarkably small.}

    \deleted{We also tested the ResNet-101 model on diffractograms simulated
    based on the statistics dataset from which the statistics for our generation
    algorithm have been extracted. This returns a $4.33$ percentage points
    higher accuracy than on the test dataset. This is due to the different
    structure types present in the test and statistics dataset yielding
    different occupation probabilities (potentially also 0) for Wyckoff
    positions and overall slightly different statistics for the synthetic
    generation algorithm.}
    \deleted{To summarize, the contributions to the gap stem from differences in the
    distribution of lattice parameters and coordinates with $6.32$ percentage
    points and from the differences between the statistics dataset and test
    dataset with $4.33$ percentage points. Added together (assuming the effects of geometry and Wyckoff position occupation probabilities are unrelated and thus additive), the contributions
    almost completely explain the gap of $12.3$ percentage points between
    training and test accuracy. The remaining part is likely due to our
    algorithm that places atoms on Wyckoff positions not reproducing the ICSD
    distribution exactly. However, the remaining difference is remarkably small.}

    \begin{figure}[!htb]
    \centering
    \includegraphics{./figures/histograms/histograms_rel.png}
    \caption{Classification \replaced{error}{performance} for each bin of a) the unit cell volume
    (conventional cell settings) and b) the number of atoms in the asymmetric
    unit. \deleted{The classification performance when testing on the ICSD is shown using
    a stacked bar plot, where the red bar indicates the classification error and
    the green bar indicates the correctly classified portion of each bin. When
    testing on the synthetic distribution, the classification error is given by
    the light blue line, the rest (between dark blue and light blue line) was correctly
    classified.} The reported classification performance is based on the
    ResNet-101 model trained on diffractograms from synthetic crystals. This
    visualization clearly shows the error rate within each bin, in contrast to
    Figure~\ref{fig:histograms}, which additionally includes the relative
    proportion of the crystals of the respective bin to the total amount of
    crystals.}
    \label{fig:histograms_rel}
    \end{figure}

    In Figure~\ref{fig:histograms_rel} we show the test classification error in
    each bin for the unit cell volume and the number of atoms in the asymmetric
    unit using the ResNet-101 model trained on diffractograms of synthetic
    crystals. The classification error is shown both for testing on
    diffractograms from synthetic crystals and on ICSD diffractograms. One can
    see that for small volumes and a small number of atoms in the asymmetric
    unit, the difference between classifying ICSD diffractograms and
    diffractograms from synthetic crystals is relatively small. As the volume
    and number of atoms in the asymmetric unit increase, the gap between the two
    errors increases, too. We already identified the uniformly sampled atom
    coordinates in the synthetic distribution as the main contributor to the gap
    in accuracy between the synthetic crystals and ICSD crystals. Therefore, it
    seems that the uniform sampling of atom coordinates works well for small 
    number of atoms in the asymmetric unit and small volumes, while the error 
    due to this sampling strategy increases slightly for higher volumes and higher
    number of atoms in the asymmetric unit.

    When looking at the distribution of crystals in the ICSD, the number of
    atoms in the asymmetric unit tends to be larger for lower-symmetry space
    groups (for example, in the triclinic crystal system) than for
    higher-symmetry space groups such as those from the cubic crystal system.
    Therefore, the increasing test error on diffractograms from ICSD crystals
    with a higher number of atoms in the asymmetric unit is especially relevant
    for these lower-symmetry space groups.
    It might be possible that a different scheme of generating atom positions in
    the unit cell (compared to the independent uniform sampling that we used)
    works better for a high number of atoms in the asymmetric unit.

    Overall, it is important to note that the distribution of ICSD crystals is (apart from a few Wyckoff position occupation probabilities which are exactly zero in the statistics dataset\footnote{Setting them to small non-zero values typically leads to the generation of rather large unit cells, as the general Wyckoff positions have high multiplicities.}) almost completely encompassed by the much larger distribution of snynthetic crystals that we used for training.
    However, due to finite training times and model capacity, a performance gap remains. This gap can be improved by using (substantially) more computing power or by narrowing the very general synthetic distribution, e.g., by using a different algorithm to generate atom positions.
    This indicates an inherent challenge in XRD classification but more generally in materials property prediction:
    Machine learning models are ultimately trained to be employed in real-world tasks, which are typically related to novel, i.e. yet unseen materials and structures.
    At the same time, the machine learning models are tested based on an IID assumption, i.e. the assumption that the distribution of training and testing data is the same.
    While not being a contradiction in the limit of infinite training data and model capacity, this becomes an (unsolvable) challenge in reality, when facing finite datasets and models.
    In our case, our model trained on a large distribution of synthetic crystal structures will likely generalize better to completely new crystal structures different from any crystal structure contained in the ICSD database.
    At the same time, it suffers from smaller ICSD test set errors, even though the ICSD distribution is contained in the synthetic data generation distribution.

>>>>

% Sharing raw powder x-ray diffraction data: Challenges and benefits https://www.researchgate.net/publication/329256004_Sharing_powder_diffraction_raw_data_Challenges_and_benefits

% The subsections of this section should be the 
% individual contributions that we get from labs
